---
title: "R Notebook"
output: html_notebook
---

In this notebook I create an animation that shows how Voronoi polygons can be created by expanding circles until they intersect. This image contains several key elements of spatial statistics: points, areas, and continuous processes, as well as the problem of change of support. To run this notebook, I use data from the data package that I use to teach spatial statistics:
```{r}
if(!require(geog4ga3)){
  remotes::install_github("paezha/geog4ga3")
  library(geog4ga3)
}
```

Load packages
```{r message=FALSE, warning=FALSE}
library(deldir)
library(geog4ga3)
library(gganimate)
library(ghibli)
library(MexBrewer)
#library(plotly)
#library(spatstat)
library(spdep)
library(tidyverse)
```
Convert Walker Lake data to simple features:
```{r}
# Function `st_as_sf()` takes a foreign object (foreign to the `sf` package) and converts it into a simple features object. If the foreign object is points, the coordinates can be named by means of the argument `coords`. 
Walker_Lake.sf <- Walker_Lake %>% 
  st_as_sf(coords = c("X", "Y"))
```

Define a bounding box to extract a subset of points:
```{r}
# Function `st_polygon()` creates an `sf` object with a polygon or polygons. In this case, we create a single polygon, a rectangle with corners given by the coordinates in the function. 
W.bbox <- st_polygon(list(rbind(c(130, 140),c(190, 140),c(190, 220),c(130, 220),c(130, 140))))
```

Extract subset of points using the bounding box:
```{r}
Walker_Lake.sf <- Walker_Lake.sf[W.bbox,]
```

Once we have an `sf` object of the points, the geometry can be used to create the Voronoi polygons:
```{r}
# The function `do.call(what, arg)` applies a function `what` to the argument `arg`. In this case, we extract the geometry of the `sf` object (i.e., the coordinates of the points) and apply the function `c()` to concatenate the coordinates to obtain a MULTIPOINT object.   
vpolygons <- do.call(c, st_geometry(Walker_Lake.sf)) %>% # The pipe operator passes the MuLTIPOINT object to function `st_voronoi()`
  st_voronoi() %>% # The output of `st_voronoi()` is a collection of geometries, which we pass to the following function for extraction.
  st_collection_extract()
```

Obtain the geometry of the polygons:
```{r}
Walker_Lake.v <- Walker_Lake.sf
Walker_Lake.v$geometry <- vpolygons[unlist(st_intersects(Walker_Lake.sf, vpolygons))] 
```

The intersection of the polygons with the box clips the polygons:
```{r warning=FALSE}
Walker_Lake.v <- Walker_Lake.v %>%
  st_intersection(W.bbox)
```

Plot the polygons:
```{r}
ggplot() + 
  geom_sf(data = Walker_Lake.v, 
          aes(fill = V), 
          color = NA)
```

Use the original points and create buffers of different sizes. Save the results for each buffer size in a slice of table labelled by `t`. Each of these slices will become a frame in the subsequent animation:
```{r warning=FALSE}

Walker_Lake.c <- data.frame()

buf <- c(1, 2, 4, 6, 8, 10, 12, 14, 25)#seq(1, 25, 4)

for(t in 1:length(buf)){
  Walker_Lake.b <- Walker_Lake.sf %>% 
    st_buffer(dist = buf[t])
  
  Walker_Lake.j <- data_frame()
  
  for(i in 1:nrow(Walker_Lake.b)){
    Walker_Lake.j <- rbind(Walker_Lake.j,  
                           st_intersection(Walker_Lake.b[i,],
                                           Walker_Lake.v$geometry[i]))
  }
  Walker_Lake.j$t <- as.character(t)
  Walker_Lake.c <- rbind(Walker_Lake.c, Walker_Lake.j)
}
```

Example of plot with polygons and buffers at t = 2:
```{r}
ggplot() + 
  geom_sf(data = Walker_Lake.c %>% 
            filter(t == "2"), 
          aes(fill = V), 
          color = NA) + 
  geom_sf(data = Walker_Lake.v, fill = NA) +
  theme_void() +
  theme(legend.position = "none") 
```

Use {gganimate} to create the animation. In this animation I use the mononoke color palette from the {ghibli} package:
```{r}
anim <- ggplot() + 
  geom_sf(data = Walker_Lake.c,
          aes(fill = V),
          color = NA) + 
  geom_sf(data = Walker_Lake.v,
          fill = NA,
          color = "black") +
  geom_sf(data = Walker_Lake.sf,
          size = 3,
          color = "black") +
  scale_fill_ghibli_c("MononokeLight", direction = -1) +
  theme_void() +
  theme(legend.position = "none")  +
  transition_states(t,
                    transition_length = 100,
                    state_length = 1.0,
                    wrap = FALSE) +
  ease_aes()

animate(anim, 
        nframes = 200,
        start_pause = 20,
        end_pause = 20)
```

Save the animation as a gif:
```{r}
anim_save("voronoi_mononoke.gif", animation = anim)
```

Recreate the animation but now using the ponyo color palette from the {ghibli} package:
```{r}
anim <- ggplot() + 
  geom_sf(data = Walker_Lake.c,
          aes(fill = V),
          color = NA) + 
  geom_sf(data = Walker_Lake.v,
          fill = NA,
          color = "black") +
  geom_sf(data = Walker_Lake.sf,
          size = 3,
          color = "black") +
  scale_fill_ghibli_c("PonyoLight", direction = 1) +
  theme_void() +
  theme(legend.position = "none")  +
  transition_states(t,
                    transition_length = 100,
                    state_length = 1.0,
                    wrap = FALSE) +
  ease_aes()

animate(anim, 
        nframes = 200,
        start_pause = 20,
        end_pause = 20)
```

Save as gif:
```{r}
anim_save("voronoi_ponyo.gif", animation = anim)
```

Use {gganimate} to create the animation. In this animation I use the Alacena color palette from the {MexBrewer} package:
```{r}
anim <- ggplot() + 
  geom_sf(data = Walker_Lake.c,
          aes(fill = V),
          color = NA) + 
  geom_sf(data = Walker_Lake.v,
          fill = NA,
          color = "black") +
  geom_sf(data = Walker_Lake.sf,
          size = 3,
          color = "black") +
  scale_fill_gradientn(colors = mex.brewer("Alacena")) +
  theme_void() +
  theme(legend.position = "none")  +
  transition_states(t,
                    transition_length = 100,
                    state_length = 1.0,
                    wrap = FALSE) +
  ease_aes()

animate(anim, 
        nframes = 200,
        start_pause = 20,
        end_pause = 20)
```

Save the animation as a gif:
```{r}
anim_save("voronoi_alacena.gif", animation = anim)
```
